Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework

### 2016-10-27 ms/ms.Rnw
More changes to intro and discussion

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits and temporal plot-scale effects -- these distinct drivers are yet to be integrated within a cohesive and predictive framework. Here we apply classic survival analysis techniques within a cross-validated hierarchical Bayesian framework to derive instantaneous mortality functions that integrate individual, species, and temporal effects for 203 tropical rain forest tree species at Barro Colorado Island (BCI) Panama. We show that mortality models that partition mortality into growth-dependent and growth-independent components, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations. Moreover, we show that the effects of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability in species mortality rates, and that temporal variation in mortality was large. Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

\dropcap{R}ates of plant mortality are known to be highly variable among individuals, among coexisting species, between forests and from year-to-year [refs]. This variation has considerable consequences for forest structure and dynamics. For example, death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools\cite{Chave:2005hz}. Furthermore it creates a gap in the canopy that can restart a successional race, during which 10's-100's more plants may die while competing for a spot in the sun. In models of forest dynamics, variation in mortality rates have been found to have a larger impact on forest structure than variation in absolute growth rates\cite{Johnson-2016}. As such, improving our ability to understand the mortality process is a priority for making accruate predictions about forest and carbon dynamics as well as water or nutrient cycles; especially in an era of rapid environmental change.

@@

The other difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of the underlying instantaneous mortality (hazard) function that accurately reflect observed patterns in mortality. Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate\cite{Kobe:1995tw,Wyckoff:2002ul}, traits\cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size\cite{Vieilledent:2010fv, Hurst:2011ky}. Generally, however, these influences have not been formalised as hazard functions\cite{Zens:2003jv}. Yet, such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function\cite{Hawkes:2000ib;Zens:2003jv,McDowell:2011dr}. This is because hazard functions explicitly define the instantaneous rate of death per unit time and consequently provide vital information on when a death is likely to occur and how various covariates affect this.

A particular challenge in developing hazard functions for plants is to integrate the observed conflicting relationships between mortality and growth within and across species. Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies provide broad support of this theory, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. By contrast, across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality rates than individuals from slower growing species\cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume). Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack\cite{Augspurger:1984wx}, and ultimately reduced mortality due to stochastic events\cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

Another major challenge is using binary observations (e.g. alive = 0; dead = 1) to infer the contribution of growth-dependent and growth-independent (i.e. stochastic events) hazards on tree mortality rates. By partitioning mortality rates into these two processes model predictive capacity may be improved and it allows us to narrow in on whether growth-dependent or growth-independent processes are the predominate cause of death in an ecosystem. For example, are tree deaths in rainforest predominately a result of competitive influences (i.e. growth-dependent processes) or are they more likely to result from growth-independent events such as treefall? More importantly, however, such partitions will allow us to examine how plant traits affect each hazard. For example, higher wood density is thought to convey higher survival by decreasing mortality aociated with stoachastic events such as those listed above. However, research has also shown that wood density may increase tolerance to low growth rates, and thus alter the growth-dependent mortality. For example, studies have found that species with high wood density are more tolerant to both shade \cite{vanGelder:2006exa}, neighbor competition \cite{Kunstler:2016km} and drought induced carbon starvation \cite{McDowell:2011dr}. Consequently, wood density may affect a trees response to both growth-dependent and growth-independent hazards. The question is, which hazard does it affect most?

@@

In this study, we reconcile both intraspecific and interspecific drivers and attempt to partition instantaneous mortality rates into growth-dependent and growth-independent hazards by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework. To derive this unified model, we compare the predictive accuracy of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig. \ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama. The simplest model (Fig. \ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (Fig. \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases. Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality decaying exponentially until an asymptote at zero is reached. The third model (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models. Specifically, this model allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (Moorcroft 2001; McDowell et al 2011). However, these alternative models have not previously been systematically compared.

In deriving the unified model, we address the following specific questions:
\begin{enumerate}\itemsep 0pt
  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
  \item What is the proportion of inter-specific variability explained by wood density?
  \item What is the proportion of total variability explained by species, census, baseline hazard and growth-dependent hazard?
  \item Does mortality rates vary substantially between censuses?
\end{enumerate}
Answering these questions involves comparing predictive accuracy of alternative models. To achieve this, we used K-fold cross-validation (Fig. \ref{fig1}D). Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations. For example, some of our models took over a two weeks to complete. We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increasing model complexity (see methods).

@@

The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards\cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}. We show that regardless of growth measure, incorporating both hazards significantly improves model predictive accuracy. This is because the additive effects of both hazards explicitly allow models to account for two independent causes of plant mortality. The growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stressors (e.g. competition, parasitization, localized herbivory). By contrast, the growth-independent hazard accounts for deaths caused by stochastic events that arise irrespective of an individuals growth rate (e.g. windthrow).

Another advantage of partitioning mortality into growth-depedent and growth-independent hazards was that we could estimate the proportion of variation each attributed. Like many other studies\cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlight the importance of light competition an important factor influencing tropical tree demographic rates. Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted varability in mortality rates (Fig. \ref{fig5}), and thus, suggests that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.

Incorporating the effect of wood density on mortality rates also improved predictive performance. Our analyses revealed that the most parsimonious combination of wood density effects was when it was attributed to only growth-independent mortality. Specifically, we found that high wood density species had lower growth-independent mortality rates relative to their low wood density counterparts. This finding corroborating the observed negative correlation between mortality and wood density reported elsewhere\cite{Kraft:2010kq,Chave:2009iy}. More importantly, however, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}.

@@

The underestimation of inter-specific variation suggests that there are other species attributes besides wood density that affect tree mortality rates. It has been suggested that a species light demand and maximum size are two major axes of inter-specific variation in tropical rainforests\cite{Poorter:2006tq}. While we did not explicitly model mortality as a function of these traits, our posthoc correlations  \ref{fig6} revealed that some inter-specific variation was indeed correlated with light demand. Specifically, we found that light demanding species (i.e. those with high gap index) were more susceptible to death due to both low growth and stochastic chance and required faster growth rates to achieve a given mortality rate relative to a species that readily recruits in shaded conditions, supporting findings by others\cite{Ruger:2011cv}. By contrast, we detected no correlations between a species maximum stem diameter and mortality rates, supporting findings by others\cite{Wright:2010fl}.

@@

Future research should also focus on examining how climate variables influence both growth-depedent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}. Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010. Unfortuantely, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3). However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by a El-Nino in 1997\textendash 1998. Droughts are known to have significant impacts on tree mortality rates in these ecosystems\cite{Condit:1995wl, Mcdowell:2013wf} Furthermore, future research should also examine how different growth measures influence predictive accuracy of mortality models. We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive accuracy, and we are unsure how other common measures such as biomass growth\cite{Chave:2005hz} will compare.

@@

Example referencing suppmat figure (Fig. \ref{figS1}).

@@

Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions. In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of observed growth rates were negative. While there are physiological processes (e.g. water depletion and bark drop) that can lead to trunk shrinkages overtime\cite{Pastur:2007kl}, it is biologically impossible for growth (i.e. the creation of new biomass) to be negative\cite{Clark:2007ve}. To ensure our mortality model was not biased by these implausible values we first developed a model to estimate \textit{``true growth''}(i.e. observed growth minus observation error). Details of this model can be found in supplementary material (REF, FigS1).

@@

For all models we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}. We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic\cite{Gelman:1992ts, Brooks:1998ju}. After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior. These models were then cross validated using 10-fold cross validation (REF).

>>>>

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits and temporal effects -- these distinct drivers are yet to be integrated within a cohesive and predictive framework. In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive instantaneous mortality functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. Specifically, we develop mortality functions that not only integrate individual, species and temporal effects, but also simultaneously estimate the relative contribution of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate. We show that functions that partition mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations. Moreover, we show that the effect of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability observed between species, and that temporal variation in mortality was large. Lastly, our analysis indicate that growth-dependent processes are the predominant contributor to rates of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests. More importantly, it highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

\dropcap{R}ates of plant mortality are known to be highly variable among individuals, coexisting species, between forests and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}. This variation has considerable consequences for forest structure and dynamics. For example, death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools\cite{Chave:2005hz}. Furthermore it creates a gap in the canopy that can restart a successional race, during which 10's-100's more plants may die while competing for a spot in the sun. In models of forest dynamics, variation in mortality rates have been found to have a larger impact on forest structure than variation in absolute growth rates\cite{Johnson-2016}. As such, improving our ability to understand the mortality process is a priority for making accruate predictions about forest and carbon dynamics as well as water or nutrient cycles; especially in an era of rapid environmental change.

@@

The other difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of the underlying instantaneous mortality (hazard) function that accurately reflect observed patterns in mortality. Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate\cite{Kobe:1995tw,Wyckoff:2002ul}, traits\cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size\cite{Vieilledent:2010fv, Hurst:2011ky}. Generally, however, these influences have not been formalised as hazard functions\cite{Zens:2003jv}. Yet, such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function\cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}. This is because hazard functions explicitly define the instantaneous rate of death per unit time and consequently provide vital information on when a death is likely to occur and how various covariates affect this.

\section*{Unifying intra- and inter-specific variation}

A particular challenge in developing hazard functions for plants is to integrate the conflicting relationships observed between mortality and growth within and across species. Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies provide broadly support this theory, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. By contrast, across species, there tends to be a strong trade-off between growth and survival, with faster growing species having higher mortality rates than slower growing species\cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume). Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack\cite{Augspurger:1984wx}, and ultimately reduced mortality due to stochastic events\cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

Another major challenge is estimating the relative contribution of growth-dependent and growth-independent hazards on an individual's overall instantaneous mortality rate. While plants can be killed by many factors, they all broadly fall into two classes. Plants can die due to insufficient carbon assimulation required for growth and repair (i.e. growth-dependent hazards), or they can die because of stochastic events that are independent of an individual's growth rate (e.g.  windfall, fire). By developing models that can simultaneously estimate growth-dependent and growth-independent hazards we can begin to quantify their relative contribution and examine how they vary by ecosystem. More importantly, such models will allow us to determine how plant traits influence each hazard. For example, wood density is thought to primarily affect growth-independent mortality by reducing stem breakage \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}. However, recent research has also indicated that higher wood density conveys increased tolerance to shade \cite{vanGelder:2006exa}, neighbor competition \cite{Kunstler:2016km} and drought induced carbon starvation \cite{McDowell:2011dr}. The question is, does wood density affect both hazards? And if so, which hazard does it affect most? Answering such questions will ultimately allow us to incorporate more biological information into existing mortality functions, which in turn, will increase model predictive capacity.

@@

In this study, we reconcile both intraspecific and interspecific drivers and attempt to partition instantaneous mortality rates into growth-dependent and growth-independent hazards by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework. To derive this unified model, we compare the predictive accuracy of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig. \ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama. The simplest model (Fig. \ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (Fig. \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases. Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality decaying exponentially until an asymptote at zero is reached. The third model (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models. Specifically, this model allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}. However, these alternative models have not previously been systematically compared.

In deriving the unified model, we address the following specific questions:
\begin{enumerate}\itemsep 0pt
  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
  \item What is the proportion of inter-specific variability explained by wood density?
  \item What is the proportion of total variability explained by species, census and growth-dependent and growth-independent hazards?
  \item Does mortality rates vary substantially between censuses?
\end{enumerate}
Answering these questions involves comparing predictive accuracy of alternative models. To achieve this, we used K-fold cross-validation (Fig. \ref{fig1}D). Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations. For example, some of our models took over a two weeks to complete. We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increasing model complexity (see methods).

@@

The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards\cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}. We show that regardless of growth measure, incorporating both hazards significantly improves model predictive accuracy. This is because the additive effects of both hazards explicitly allow models to account for two independent causes of plant mortality. The growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stressors (e.g. competition, parasitization, localized herbivory). By contrast, the growth-independent hazard (i.e. baseline hazard) accounts for deaths caused by stochastic events that arise irrespective of an individuals growth rate (e.g. windthrow).

Another advantage of partitioning mortality into growth-dependent and growth-independent hazards was that we could estimate the proportion of variation attributed to each. Like many other studies\cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlight the importance of light competition in influencing tropical tree demographic rates. Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted variability in mortality rates (Fig. \ref{fig5}), indicating that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.

Incorporating the effect of wood density on mortality rates also improved predictive performance. Our analyses revealed that the most parsimonious combination of wood density effects was when it was attributed to only growth-independent mortality. Specifically, we found that high wood density species had lower growth-independent mortality rates relative to their low wood density counterparts. This finding corroborates the observed negative correlation observed between mortality and wood density reported elsewhere\cite{Kraft:2010kq,Chave:2009iy}. More importantly, however, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}.

@@

The high unexplained inter-specific variation suggests that there are other species traits besides wood density that affect tree mortality rates. It has been suggested that a species light demand and maximum size are two major axes of inter-specific variation in tropical rainforests\cite{Poorter:2006tq}. While we did not explicitly model mortality as a function of these traits, our posthoc correlations \ref{fig6} revealed that some inter-specific variation was indeed correlated with light demand. Specifically, we found that light demanding species (i.e. those with high gap index) were more susceptible to death due to both low growth and stochastic chance and required faster growth rates to achieve a given mortality rate relative to a species that readily recruits in shaded conditions, supporting findings by others\cite{Ruger:2011cv}. By contrast, we detected no correlations between a species maximum stem diameter and mortality rates, supporting findings by others\cite{Wright:2010fl}.

@@

Furthermore, future research should also focus on examining how climate variables influence both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}. Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010. Unfortuantely, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3). However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by a El-Nino in 1997\textendash 1998. Droughts are known to have significant impacts on tree mortality rates in these ecosystems\cite{Condit:1995wl, Mcdowell:2013wf}.

The length between censuses should also be examined to determine its affect on the estimation of growth-dependent and growth-independent hazards. Large census intervals are likely to lead to underestimation of growth-dependent mortality, and consequently, an overestimation of growth-independent mortality. This is because if a census interval is too wide, it will fail to account for deaths attributed to rapid declines in growth caused by senescence (especially for short-lived plants) or events such as drought. Because our analysis estimated an individual's past growth from 5 year census intervals, and because it included short-lived species (i.e. those with high annual mortality rates; see Fig. \ref{fig4}) and a drought period between 1995 and 2000, our estimates of growth-dependent and growth-independent hazards are likely to be biased.

Lastly, research should also examine how different growth measures influence predictive accuracy of mortality models. We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive accuracy, and we are unsure how other common measures such as biomass growth\cite{Chave:2005hz} will compare.

@@

@@

Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions. In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of observed growth rates were negative. While there are physiological processes (e.g. water depletion and bark drop) that can lead to trunk shrinkages overtime\cite{Pastur:2007kl}, it is biologically impossible for growth (i.e. the creation of new biomass) to be negative\cite{Clark:2007ve}. To ensure our mortality model was not biased by these implausible values we first developed a model to estimate \textit{``true growth''}(i.e. observed growth minus observation error). Details of this model can be found in supplementary material (FigS1).

@@

For all models we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}. We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic\cite{Gelman:1992ts, Brooks:1998ju}. After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior. These models were then cross validated using 10-fold cross validation.
